Variance reduction for Markov chains via martingale representations

### 2020-05-02 aap/Revision/vr-mcmc-revision-new-new.tex
update

Now, from the H\"older inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
\begin{align*}
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
 & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
\end{align*}
Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1}^x)\right),$
we get
\[
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
\]
where we denote
\[
\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right],\quad u>0.
\]
Note that due to our assumptios, $\mu(x) = \mu_0(x) + \Delta(x)$ with $\mu^{\prime}_0(x) < -m$. Thus by applying Holder inequality to $\zeta_{l,j}(u)$, we obtain
\begin{equation}
\label{eq:zeta}
\zeta_{l,j}(u) \leq \left(\mathsf{E}\left[\rme^{-2u\gamma\sum_{k=l}^{j}\mu_0'(X^x_{k-1})}\right]\right)^{1/2}\left(\mathsf{E}\left[\rme^{-2u\gamma\sum_{k=l}^{j}\Delta'(X^x_{k-1})}\right]\right)^{1/2}
\end{equation}
Note that $\left|\Delta^{\prime}(x)\right| \leq M^{\prime}$, hence
\[
\left(\mathsf{E}\left[\rme^{-2u\gamma\sum_{k=l}^{j}\Delta'(X^x_{k-1})}\right]\right)^{1/2} \leq \mathsf{E}\rme^{2u(j-l)M^{\prime}}
\]
To bound the first exponential moment in \ref{eq:zeta}, we apply
Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix}. It implies that
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l)\gamma\left(4u^2\kappa^{-2} \|\mu^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + C_1u\sqrt{\gamma} -  u\pi(\mu')\right)\right\}, \quad \gamma(j-l) \geq C_0 \\
\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
\end{cases}
\end{align*}
with $C_0 = \left(\frac{4}{m+L} + \kappa^{-1}\right)$, and constants $C(x)$ and $C_1$ given in Lemma~\ref{lem:exp_bound} . Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
\[
\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=\frac{1}{2}\int\mu^{2}(x)\pi(x)\,dx>0
\]
Hence the following bound follows
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{-\gamma(j-l)\alpha\right\}, \quad \gamma(j-l) \geq C_0 \\
\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
\end{cases}
\end{align*}
with the quantity $ \alpha = \min(...,...) > 0$. Hence we obtain
\begin{align*}
\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + C(x)\sum_{p=j+1}^{\infty} \rme^{-\gamma(p-j)\alpha}
\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
\end{align*}
and
\begin{align*}
\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
\end{align*}
Thus the final bound follows:
\begin{align*}
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K C(x)
\end{align*}
 with $C_K$  depending neither on $q$ nor on $\gamma$. The proof is completed.
\end{proof}

@@

In all the sequel, without loss of generality, we assume that $\nabla U(0)=0$ and we set
\begin{equation}
\label{eq:definition-V}
V(x) = 1 + \|x\|^2 \eqsp.
\end{equation}
\begin{lem}
  \label{lem:quadratic_behaviour}
  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ \| x \| \geq K_2$ $\ps{\nabla U(x)}{x} \geq (m/2) \| x \|^2$.
\end{lem}
\begin{proof}
  Using {\bf (H1)} and {\bf (H2)}, we have for any $x \in \rset^d$, $\| x \|\geq K_1$,
  \begin{align*}
    \ps{\nabla U(x)}{x}
    &= \int_{0}^{K_1/\| x \|} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/\| x \|} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
    & \geq m \|x \|^2 \{1- K_1 (1 +L/m)   / \| x \| \}\eqsp,
  \end{align*}
which proves the first statement. The second statement is obvious.
\end{proof}
\begin{lem}
\label{lem:drift}
 Assume {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
 Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = m/2L^2$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
 \begin{equation}
\label{eq:lyapunov}
Q_{\gamma}V(x) \leq \rho^{\gamma} V(x) + \gamma C
\end{equation}
with   $\rho = \exp{\left(-\frac{m}{2}\right)}$, $C = 2K_2^2(L+m) + d + m$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $m$ from {\bf (H2)}.
\end{lem}

@@

which means that the Markov kernel $\P$ is a strict contraction in Wasserstein metric. This allows to use \cite[Proposition 1]{joulin2010}, stating that

@@

\left|\sum_{k=N+1}^{N+n}\PE_x f(Y_k) - n\pi_{\gamma}(f)\right| \leq \frac{2\left(1-\frac{\kappa}{2}\right)^{N+1}}{\kappa}\|f\|_{\text{Lip}}\int_{\rset}|x-y|\,\pi_{\gamma}(dy) \leq \frac{m+L}{mL}\|f\|_{\text{Lip}}S(x)

@@

and \cite[Corollary 7]{dm2016HighDim}, that is

>>>>

Now, from the H\"older inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
\begin{align*}
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
 & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
\end{align*}
Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1}^x)\right),$
we get
\[
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
\]
where we denote
\[
\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right],\quad u>0.
\]
Note that due to our assumptios, $\mu(x) = \mu_0(x) + \Delta(x)$ with $\mu^{\prime}_0(x) < -m$. Note that $\left|\Delta^{\prime}(x)\right| \leq M^{\prime}$, hence
\begin{equation}
\label{eq:zeta}
\zeta_{l,j}(u) \leq \mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu_0'(X^x_{k-1})}\right]\rme^{u\gamma(j-l+1)M^{\prime}}
\end{equation}
To bound the first exponential moment in \ref{eq:zeta}, we apply
Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix}. It implies that
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l+1)\gamma\left(4u^2\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + C_1u\sqrt{\gamma} -  u\pi_{\mu_0}(\mu_0')\right)\right\}, \quad \gamma(j-l+1) \geq C_0 \\
\exp\left\{uC_0(\max\limits_{y}|\mu_0'(y)| + M^{\prime})\right\}, \quad 0 \leq \gamma(j-l+1) \leq C_0	
\end{cases}
\end{align*}
with $C_0 = \left(\frac{4}{m+L} + \kappa^{-1}\right)$, measure $\pi_{\mu_{0}}$ and constants $C(x)$ and $C_1$ given in Lemma~\ref{lem:exp_bound} . Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
\[
\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=\frac{1}{2}\int\mu^{2}(x)\pi(x)\,dx>0
\]
Hence the following bound follows
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{-\gamma(j-l)\alpha\right\}, \quad \gamma(j-l) \geq C_0 \\
\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
\end{cases}
\end{align*}
with the quantity $ \alpha = \min(...,...) > 0$. Hence we obtain
\begin{align*}
\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + C(x)\sum_{p=j+1}^{\infty} \rme^{-\gamma(p-j)\alpha}
\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
\end{align*}
and
\begin{align*}
\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
\end{align*}
Thus the final bound follows:
\begin{align*}
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K C(x)
\end{align*}
 with $C_K$  depending neither on $q$ nor on $\gamma$. The proof is completed.
\end{proof}

@@

In all the sequel, without loss of generality, we assume that $\nabla U(0)=0$ and we set
\begin{equation}
\label{eq:definition-V}
V(x) = 1 + \|x\|^2 \eqsp.
\end{equation}
\begin{lem}
  \label{lem:quadratic_behaviour}
  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ \| x \| \geq K_2$ $\ps{\nabla U(x)}{x} \geq (m/2) \| x \|^2$.
\end{lem}
\begin{proof}
  Using {\bf (H1)} and {\bf (H2)}, we have for any $x \in \rset^d$, $\| x \|\geq K_1$,
  \begin{align*}
    \ps{\nabla U(x)}{x}
    &= \int_{0}^{K_1/\| x \|} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/\| x \|} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
    & \geq m \|x \|^2 \{1- K_1 (1 +L/m)   / \| x \| \}\eqsp,
  \end{align*}
Now the result follows given that $K_1 (1 +L/m)   / \| x \| \geq \frac{1}{2}$.
\end{proof}
\begin{lem}
\label{lem:drift}
 Assume {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
 Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = m/2L^2$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
 \begin{equation}
\label{eq:lyapunov}
Q_{\gamma}V(x) \leq \rho^{\gamma} V(x) + \gamma C
\end{equation}
with   $\rho = \exp{\left(-\frac{m}{2}\right)}$, $C = 2K_2^2(L+m) + d + m$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $m$ from {\bf (H2)}.
\end{lem}

@@

which means that the Markov kernel $\P$ is a strict contraction in the Wasserstein metric. This allows to use \cite[Proposition 1]{joulin2010}, stating that

@@

\left|\sum_{k=N+1}^{N+n}\PE_x f(Y_k) - n\pi_{\gamma}(f)\right| \leq \frac{2\left(1-\frac{\kappa\gamma}{2}\right)^{N+1}}{\kappa\gamma}\|f\|_{\text{Lip}}\int_{\rset}|x-y|\,\pi_{\gamma}(dy) \leq \frac{m+L}{mL\gamma}\|f\|_{\text{Lip}}S(x)

@@

with $S(x) = \int_{\rset}|x-y|\,\pi_{\gamma}(dy)$. Then by \cite[Corollary 7]{dm2016HighDim}, 
